---
title: "Getting Started with gtfsaccess"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with gtfsaccess}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(gtfsaccess)
```

# Introduction

Public transport planners often want to answer a simple but important
question: **do people have enough transit service near where they live?**
Answering this requires combining three things that normally live in
different files and different formats: GTFS transit schedules, a map of
transit stops, and gridded population data.

**gtfsaccess** brings these together. Give it a GTFS feed and a
population raster, and it will locate every transit stop, draw a
"catchment area" around each stop, count how many people live in that
catchment, count how many departures serve that stop, and combine
these into a single accessibility index you can summarise, classify,
and map.

Three S3 classes are used in this vignette:

- **`gtfs_stops`** - stop locations read from a GTFS feed
- **`gtfs_accessibility`** - accessibility results for each stop's catchment
- **`gtfs_district_accessibility`** - accessibility results aggregated to district level

Each supports `print()` and `plot()`; `gtfs_stops` and
`gtfs_accessibility` also support `summary()`. All plotting including
choosing colour classes and drawing legends is handled inside these
methods, so the code below only ever calls the package's public
functions.

A note on the maps in this vignette: they are intentionally exploratory
(no north arrow or scale bar), since the goal here is to compare
relative accessibility across catchments and districts rather than to
produce a print-ready reference map.

## Typical Workflow

```
GTFS ZIP file + study area boundary
              |
              v
      read_gtfs_stops()          ->  gtfs_stops object
              |
              v
      create_service_areas()     ->  Voronoi catchment polygons
              |
              v
      calculate_accessibility()  ->  gtfs_accessibility object
              |
              v
      classify_accessibility()   ->  gtfs_accessibility object with classes (optional)
              |
              v
      aggregate_accessibility()  ->  gtfs_district_accessibility object (optional)
```

Every step below shows exactly one of these functions in action, using
small example data bundled with the package, so every chunk actually
runs.

# Step 1: Reading GTFS Stops

The starting point is a GTFS ZIP file, the standard format transit
agencies use to publish their schedules and stop locations. `gtfsaccess`
ships a small example feed so you can follow along without downloading
anything.

```{r}
gtfs_zip <- system.file(
  "extdata", "example_gtfs.zip",
  package = "gtfsaccess"
)

stops <- read_gtfs_stops(gtfs_zip)

print(stops)

summary(stops)
```

```{r, fig.width=5, fig.height=5}
plot(stops)
```

`summary()` also reports the coordinate reference system: stops are
automatically reprojected to EPSG:3035, since later steps need
distances and areas in real-world units rather than degrees of
latitude/longitude.

## How many stops are there, and how densely spread out are they?

```{r}
boundary_file <- system.file(
  "extdata", "example_boundary.geojson",
  package = "gtfsaccess"
)

stop_statistics(stops, boundary_file)
```

## Defining each stop's service area

Catchments are drawn as Voronoi polygons. Each stop "claims" all the
space closer to it than to any other stop, clipped to the study area
boundary.

```{r, fig.width=5, fig.height=5}
service_areas <- create_service_areas(
  stops,
  boundary_file,
  plot = TRUE
)
```

# Step 2: Calculating Accessibility

## The formula

For each catchment $i$:

$$
\text{accessibility_index}_i =
\frac{\text{departures}_i}{\text{population_in_catchment}_i} *100
$$

Catchments with zero population are assigned `NA`, since "service per 100
residents" is undefined when there is no one there to be served.

## What the index means

The index is best interpreted as **the number of scheduled departures available per 100 residents within a stop's service catchment**. It does not account for where those
departures actually go (see *Limitations* below). The measure is more precisely 
a per-capita transit supply indicator. Throughout every map
in this vignette, **darker/more intense colour always means a higher
index**.

- **Low accessibility** → few departures relative to the surrounding
  population, that means many residents sharing the available transit service, infrequent service, or both.
- **High accessibility** → many departures relative to the nearby
  population that means frequent service, a sparse catchment, or both.

Because a single index value is compatible with more than one
underlying cause, `population_in_catchment` and `departures` are kept
in the output so either factor can be checked directly.

## Computing and viewing it

```{r}
population_file <- system.file(
  "extdata", "example_population.tif",
  package = "gtfsaccess"
)

population <- terra::rast(population_file)

accessibility <- calculate_accessibility(
  service_areas,
  population,
  gtfs_zip
)

print(accessibility)

summary(accessibility)
```

The mean is substantially larger than the median and a small number of catchments have exceptionally high accessibility values, indicating a strongly right-skewed distribution.

The raw index, plotted directly:

```{r, fig.width=5, fig.height=5}
plot(accessibility)
```

`plot()` classes catchments into quantile bins rather than equal-width
ones, since accessibility indices are typically right-skewed (a small
number of catchments are often around a single dominant hub have very
high values). Quantile classing spreads catchments evenly across
colour bins regardless of that skew.

## Turning numbers into categories

`classify_accessibility()` sorts catchments into named, ordered
categories using Jenks natural breaks by default:

```{r}
accessibility <- classify_accessibility(accessibility)

table(accessibility$accessibility$accessibility_class)
```

Plotting the same object with `classified = TRUE` now uses those
stored classes:

```{r, fig.width=5, fig.height=5}
plot(accessibility, classified = TRUE)
```

**Why is almost everything "Very Low"?** This reflects the underlying
data, not a plotting artefact: accessibility here is concentrated
around one or two major hub stops with disproportionately high
departures-per-100-resident values, while the rest of the network sits far
below them  in this example data, likely driven by a small number of
catchments with small population but non-trivial service (a
known sensitivity of the raw formula; see *Limitations*).

Plotted directly, these extreme values dominate the classification and reduce the visual contrast among the remaining catchments.


```{r}
accessibility_log <- accessibility

accessibility_log$accessibility$accessibility_index <-
  log1p(accessibility_log$accessibility$accessibility_index)

accessibility_log <- classify_accessibility(accessibility_log)
table(accessibility_log$accessibility$accessibility_class)
```

To improve visualization, we apply a logarithmic transformation to the accessibility index before classification. The transformation compresses the largest values while preserving the ordering of catchments, producing a more balanced classification without changing the underlying accessibility measure computed by `calculate_accessibility()`.

```{r, fig.width=5, fig.height=5}
plot(accessibility_log, classified = TRUE)
```


Other classification schemes are available too, for example, tertiles
via quantiles, which guarantee roughly equal numbers of catchments per
class:

```{r}
accessibility_quantile <- classify_accessibility(
  accessibility,
  n_classes = 4,
  method = "quantile"
)

table(accessibility_quantile$accessibility$accessibility_class)
```

```{r, fig.width=5, fig.height=5}
plot(accessibility_quantile, classified = TRUE)
```

Using quantiles forces approximately the same number of catchments into each class. Unlike Jenks natural breaks, which highlights natural groupings in the data, quantiles emphasize relative ranking. The resulting map spreads colours more evenly across the study area, making spatial patterns easier to compare, although class boundaries may no longer correspond to meaningful gaps in the accessibility values.

## Extracting summary statistics directly

```{r}
accessibility_statistics(accessibility)
```

# Step 3: Aggregating to Districts

`aggregate_accessibility()` rolls catchment-level results up to
district polygons, returning a `gtfs_district_accessibility` object
that supports the same `print()`/`plot()` pattern as above:

```{r}
districts_file <- system.file(
  "extdata", "example_districts.geojson",
  package = "gtfsaccess"
)

district_accessibility <- aggregate_accessibility(
  accessibility,
  districts_file
)

print(district_accessibility)
```

The raw, quantile-classed view:

```{r, fig.width=5, fig.height=5}
plot(district_accessibility)
```

The full underlying table including `total_population` and
`total_departures` per district is available via
`district_accessibility$accessibility`:

```{r}
district_accessibility$accessibility
```

# Summary of Methods Demonstrated

| Class                          | Standard methods            | Package-specific methods                        |
|----------------------------------|------------------------------|---------------------------------------------------|
| `gtfs_stops`                    | `print()`, `summary()`, `plot()` | `stop_statistics()`, `create_service_areas()`     |
| `gtfs_accessibility`            | `print()`, `summary()`, `plot()`, `plot(classified = TRUE)` | `accessibility_statistics()`, `aggregate_accessibility()`, `classify_accessibility()` |
| `gtfs_district_accessibility`   | `print()`, `plot()`, `plot(classified = TRUE)` | - |

# Limitations and Future Extensions

The accessibility index used here - departures per 100 residents within a
Voronoi catchment is a deliberately simple proxy for transit supply.
It does not account for *where* those departures go (destination
accessibility), time-of-day service variation, or the fact that walk
catchments in reality don't have hard Voronoi edges. It is also
sensitive to catchments with very small populations, where dividing by
a small denominator can produce disproportionately large index values exactly the kind of skew visible in the maps above. Future extensions
could incorporate destination-weighted accessibility, headway
regularity, distance-decayed catchments, or a floor on catchment
population before computing the index.


# LLM Usage

## LLM tool used:
Claude Sonnet 4.6 (via claude.ai) and ChatGPT-5 (via chatgpt.com).

## Purpose:
The LLMs were used throughout package development as programming, debugging, and writing assistants. Their contributions included:

- Debugging implementation issues, including Voronoi service area generation and stop matching.
- Assisting with the implementation and refinement of helper functions.
- Suggesting improvements to plotting methods, including colour schemes, legends, labels, titles, and map layout.
- Drafting and refining function documentation, examples, and help pages.
- Assisting with vignette writing and restructuring.
- Suggesting edge cases, validation checks, and code improvements.
- Explaining R package development practices and assisting with debugging `devtools::check()` issues.

## Percentage of code generated by LLM:

The estimated level of LLM assistance for the main source files is shown below.

| File | Estimated LLM contribution |                 Notes                    |
|------|:---------------------------:|------------------------------------------|
| `classes.R` | 20-30% | S3 class definitions and object structure were designed and implemented by the author. LLM assistance was limited to examples and code review. |
| `gtfs_stops.R` | 20-30% | Core implementation by the author. LLM assisted with documentation, examples, debugging to match each Voronoi cell to its generating stop by containment and edge cases. |
| `gtfs_accessibility.R` | 30-40% | Accessibility computation and overall implementation were primarily developed by the author. LLM assisted with departure counting, aggregation, documentation, examples, and debugging. |
| `methods.R` | 60-70% | Extensive LLM assistance with plotting methods, legends, colour scales, labels, titles, layout, and visualization improvements. |
| Documentation (`.Rd`) | 70-80% | LLM extensively assisted in drafting and refining help pages, parameter descriptions, examples, and return values. |
| Vignette | 60-70% | Workflow of the document and methodology for analysis were developed by the author. LLM assisted with writing, restructuring, examples, and revisions. |
